GSE158422: DNA methylation in Lung squamous cell carcinoma - GMEA using fry()

### 2022-11-14 GSE158433/GSE158422.Rmd
reorganising analysis: gene and geneset Rmd reports

The purpose of this analysis is to apply different enrichment analysis approaches,
and see which approach is most parsimonious with gene expressoin differences.

We will specifically be testing the GMEA approach versus existing approaches,
namely GOseq and mitch.

This study (GSE158433) is a super-series contains matched RNA-seq (GSE158420) and 
Infinium EPIC array (GSE158422) data for matched control and cancer tissues.

In this script, we will be processing the EPIC DNA methylation data and performing the
statistical analysis in limma and saving the results for downstream enrichment analysis.

>>>>

This work is to conduct the methylation data processing and differential analysis of GSE158422, which is matched
lung squamous cell carcinoma (LUSC) and control tissue.

The data is going to be saved as an Rdata file and then used for enrichment analysis in a subsequent Rmd script.

### 2022-11-14 GSE158433/GSE158422_genesetlevel.Rmd
reorganising analysis: gene and geneset Rmd reports

Here I will use fry to perform GMEA analysis.

@@

  library("tictoc")
  library("kableExtra")

@@

>>>>

There are many ways to perform enrichment analysis of Infinium methylation data.
The purpose of this work is to determine which one of these methods perform "best".
Best performance is a function of specificity, sensitivity and biological consistency.

* Specificity can be defined as the ability to control false discovery rate appropriately.
This will be evaluated by analyzing a randomised data set.

* Sensitivity is defined as the ability to identify genesets undergoing differentially methylation.
This can be evaluated if we begin with a large study with many samples, analyse it and then repeat
the process with fewer samples.
The more sensitive methods will be able to identify differentially methylated pathways with
a smaller sample size.

* As methylation is known to be a characteristic of silenced genes, we can use this feature to examine
whether the methylation pathway enrichment analysis is congruent with gene expression data, as this
dataset has matched methylation and gene expression profiles for all samples.

In addition I will be measuring at the execution time for these approaches.

The approaches that I will be examining:

* GOmeth: an approach to filter differential methylated probes followed by over-representation testing.

* ebGSEA: global test to rank genes followed by Wilcoxon Test or Known Population Median Test

* methylGSA: mRRA robust rank aggregation, mGLM

* Fry test on probes belonging to a gene, followed by Camera PR based on significance/effectsize?

* Median of probe intensity for each gene in each sample followed by Fry test on pathways.

@@

  library("tictoc")
  library("kableExtra")
  library("mitch")
  library("fgsea")

@@

## Pathway analysis

Simple GSEA using FGSEA (gene permutation) and mitch (rank-ANOVA).

```{r,fgsea}

#download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
#unzip("ReactomePathways.gmt.zip")
genesets <- gmt_import("ReactomePathways.gmt")

stat1 <- sign(as.numeric(factor(fry_res1$Direction))-1.5)  * -log10(fry_res1$PValue)
stat1[is.na(stat1)] <- 0
names(stat1) <- rownames(fry_res1)
fres1 <- fgseaMultilevel(pathways=genesets,stats=stat1)
fres1 <- fres1[order(fres1$pval),]

head(fres1,20) %>%
  kbl(caption="WG fry + fgsea Reactome results") %>%
  kable_styling(full_width=FALSE)

stat2 <- sign(as.numeric(factor(fry_res2$Direction))-1.5)  * -log10(fry_res2$PValue)
stat2[is.na(stat2)] <- 0
names(stat2) <- rownames(fry_res2)
fres2 <- fgseaMultilevel(pathways=genesets,stats=stat2)
fres2 <- fres2[order(fres2$pval),]
head(fres2,20)  %>%
  kbl(caption="Promoter fry + fgsea Reactome results") %>%
  kable_styling(full_width=FALSE)

```{r,mitch}

stat1 <- data.frame(stat1)
mres1 <- mitch_calc(stat1, genesets, priority="effect")
head(mres1$enrichment_result,20) %>%
  kbl(caption="WG fry + mitch Reactome results") %>%
  kable_styling(full_width=FALSE)

stat2 <- data.frame(stat2)
mres2 <- mitch_calc(stat2, genesets, priority="effect")
head(mres2$enrichment_result,20) %>%
  kbl(caption="Promoter fry + mitch Reactome results") %>%
  kable_styling(full_width=FALSE)
